Lossy compression of DNA sequencing quality data

### 2016-05-26 bmc_manuscript/manuscript.tex
BibTex: added original reference for quality values; fixed some formatting; added publisher location manuscript: cite original reference for quality values and edited related text; better describe and name third loss function based on my understanding

Quality values comprise a standard component of \textsc{fastq}
files~\cite{Cock:2010ve}, a very common format for sequence read
data. At the level of the sequence read, the probability of error for
each base-call is typically represented by a \textsc{phred} quality
value, which is defined as $Q =
-10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
technology, these quality values can range from 0 to 93, and are
represented with the \textsc{ascii} characters 33 to 126 (with some
offset). There is a single quality value per base-call for Illumina
sequence reads.

Quality values can be used throughout bioinformatics pipelines. Among
the most fundamental uses of sequence quality values is as part of the
quality assessment and quality control (\textsc{qa/qc}) processes
prior to subsequent analysis steps. Quality control based on quality
values generally includes two operations: \textit{i}.~filtering, which
removes reads that on the whole do not meet arbitrary quality
standards, thus reducing the total number of reads; and
\textit{ii}.~trimming of low-quality base-calls from reads, which
reduces the length of reads trimmed. Quality values can be used by
genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment programs, such as Bowtie2~\cite{Langmead:2012rw}, use
quality values to weight mismatches between read and reference
sequences. Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values~\cite[e.g.,][]{McKenna:2010bh},
and identified \textsc{snp}s with high-quality calls are deemed more
reliable than those with low-quality calls, particularly in
low-coverage regions.

@@

As a measure of compression effectiveness, which reflects the sum of
fixed and marginal costs, we use bits/base-call, defined as the size
of the compressed representation of quality values (in bits) divided
by the number of quality values represented. As a measure of
information loss we use mean squared error (\textsc{mse}) as a loss
function, and define it as $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$,
where $n$ is the number of quality values, $Q_i'$ is the
compressed/decompressed quality value, and $Q_i$ is the original
quality value associated with position $i$.

>>>>

Advantageous use of quality values has been long
recognize~\cite{Bonfield:1995kq}, and quality values can be used
throughout bioinformatics pipelines.  Quality values comprise a
standard component of \textsc{fastq} files~\cite{Cock:2010ve}, a very
common format for sequence read data. At the level of the sequence
read, the probability of error for each base-call is typically
represented by a \textsc{phred} quality value, which is defined as $Q
= -10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
technology, these quality values can range from 0 to 93, and are
represented with the \textsc{ascii} characters 33 to 126 (with some
offset). There is a single quality value per base-call for Illumina
sequence reads.

Among the most fundamental uses of sequence quality values is as part
of the quality assessment and quality control (\textsc{qa/qc})
processes prior to subsequent analysis steps. Quality control based on
quality values generally includes two operations:
\textit{i}.~filtering, which removes reads that on the whole do not
meet arbitrary quality standards, thus reducing the total number of
reads; and \textit{ii}.~trimming of low-quality base-calls from reads,
which reduces the length of reads trimmed. Quality values can be used
by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment programs, such as Bowtie2~\cite{Langmead:2012rw}, use
quality values to weight mismatches between read and reference
sequences. Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values~\cite[e.g.,][]{McKenna:2010bh},
and identified \textsc{snp}s with high-quality calls are deemed more
reliable than those with low-quality calls, particularly in
low-coverage regions.

@@

As a measure of compression effectiveness, which reflects the sum of
fixed and marginal costs, we use bits/base-call, defined as the size
of the compressed representation of quality values (in bits) divided
by the number of quality values represented. As measures of
information loss we use three loss functions following previous
work~\cite{Malysa01102015}: mean squared error (\textsc{mse}), defined
as $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$; mean absolute deviation
(\textsc{mad}), defined as $\frac{1}{n}\sum_{i=1}^{n}{|Q_i'-Q_i|}$;
and mean of the values derived from the Cauchy--Lorentz distribution
(mean Cauchy--Lorentz loss), defined as
$\frac{1}{n}\sum_{i=1}^{n}{log_{2}(1 + |Q_i'-Q_i|)}$, where $n$ is the
number of quality values, $Q_i'$ is the compressed/decompressed
quality value, and $Q_i$ is the original quality value associated with
position $i$.
